Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-02 intro.tex
adding experimental methods/collection section and updating

This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance but believe the results apply to a much broader class of time series.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  We study three
examples---a simple microkernel and two complex programs: one from the
SPEC 2006CPU benchmark suite, and one from LAPACK---running on an Intel i7-based machine.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle), then compare that to the prediction accuracy attainable for
that trace using a series of deterministic models.

>>>>

{\color{red} ARIMA is not deterministic due to MA so we need to change this, actually need to change this entire paragraph} This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance but believe the results apply to a much broader class of time series.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  {\color{red} Probably remove this and move as it is in the experimental methods or point to that section for more details}We study three
examples---a simple microkernel and two complex programs: one from the
SPEC 2006CPU benchmark suite, and one from LAPACK---running on an Intel i7-based machine.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle), then compare that to the prediction accuracy attainable for
that trace using a series of deterministic models.

### 2014-03-02 methods.tex
adding experimental methods/collection section and updating

\subsection{Time Series Collection}
 Experimental methods (how we collect the time series and what the times series are
This should be HPM PAPI, which programs we model

\subsection{The Programs}
Include traces of full \col \svd and \gcc
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/colFullTS}
    \caption{\col Time Series}
    \label{fig:col-ts}
  \end{subfigure}%
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
    \caption{\gcc Time Series}
    \label{fig:gcc-ts}
  \end{subfigure}
    \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/SVD1RegimesColored}
    \caption{\svd Time Series with Regime Coloring}
    \label{fig:svd-ts-colored}
  \end{subfigure}
  \caption{[[Maybe this should be just a picture of each time series and then have another figure with color after explaining heuristically the reigmes. In (a) the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
    \end{figure}

\svd brings up a very interesting point: systems change over time. While it is clear in Figure \ref{fig:sample-ts} (a) and (b) that a single system is being observed, it also appears \emph{visually} that the complexity of these two time series is consistent over time, i.e., while they are both complicated they don't appear to increase or decrease in complexity over the range of the program. \svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually. Figure \ref{fig:svd-ts-colored} is a single trace of \svd from start to finish, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series is visually very different from section to section. This is caused by the code of \svd moving between different subroutines. We call these changes in \svd dynamics  \emph{\svd regimes} and we have colored each of the six regimes different colors in Figure \ref{fig:svd-ts-colored}. The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift. In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}. The purpose of this paper is not to rigorously explore regime detection but to explore quantifying complexity of a time series. As such, making this split simply serves as providing 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore. For notational convience we refer to these signals as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd. The regimes are labeled from left to right.

>>>>

 \begin{enumerate}
 \item  Experimental methods: (how we collect the time series and what the times series areThis should be HPM PAPI, which programs we model
\item programs 
\subitem description of \col
\subitem description of \gcc
\subitem description of \svd 
\subitem description of svd regimes
 \end{enumerate}

\subsection{Time Series Collection}

The time-series data for these experiments was collected on an Intel Core\textsuperscript{\textregistered} i7-2600 running the 2.6.38-8 Linux
kernel.  This Nehalem chip has eight cores running at 3.40GHz and a cache size
of 8192 KB.  We studied three example programs---a simple microkernel(\col) and two complex programs: one from the
SPEC 2006CPU benchmark suite(\gcc), and one from LAPACK(\svd), aggregating and measuring the processor load---instructions per cycle (ipc)---at 100,000-instruction intervals.  To record these measurements,
we used the {\tt libpfm4}, via PAPI
\footnote{Performance Application Programming Interface}\cite{papi-website}, which we have instrumented to stop program execution every 100,00-instructions and read the contents of the CPU's onboard hardware performance monitors (HPMs). HPMs are specialty registers built for this purpose. Obviously using a system to measure itself can cause interference but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way. For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see 
\cite{zach-IDA10,mytkowicz09,todd-phd}.

\subsection{The Programs}
We study three example programs---a simple microkernel(\col) and two complex programs: one from the
SPEC 2006CPU benchmark suite(\gcc), and one from LAPACK(\svd).
\subsubsection{\col}
\col is a simple three-line C program that repeatedly initializes the top half of matrix in column-major order. While this is a very simple program it has been shown to exhibit very complicated behavior. In fact, in \cite{chaos} it was shown 
\subsubsection{\gcc}

\subsubsection{\svd}
\svd brings up a very interesting point: systems change over time. While it is clear in Figure \ref{fig:sample-ts} (a) and (b) that a single system is being observed, it also appears \emph{visually} that the complexity of these two time series is consistent over time, i.e., while they are both complicated they don't appear to increase or decrease in complexity over the range of the program. \svd (seen in Figure \ref{fig:svd-ts-colored}) is different however, at least visually. Figure \ref{fig:svd-ts-colored} is a single trace of \svd from start to finish, however as time progresses it is clear that something drastically changes in the underlying generating process and the structure of the time series is visually very different from section to section. This is caused by the code of \svd moving between different subroutines. We call these changes in \svd dynamics  \emph{\svd regimes} and we have colored each of the six regimes different colors in Figure \ref{fig:svd-ts-colored}. The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift. In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}. The purpose of this paper is not to rigorously explore regime detection but to explore quantifying complexity of a time series. As such, making this split simply serves as providing 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore. For notational convience we refer to these signals as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd. The regimes are labeled from left to right.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{\textwidth}
  \centering
    \includegraphics[width=0.75\textwidth]{figs/colFullTS}
    \caption{\col Time Series}
    \label{fig:col-ts}
  \end{subfigure}%
  \\

  \begin{subfigure}[t]{\textwidth}
  \centering
    \includegraphics[width=0.75\textwidth]{figs/gccfullts}
    \caption{\gcc Time Series}
    \label{fig:gcc-ts}
  \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[width=0.75\textwidth]{figs/SVD1RegimesColored}
    \caption{\svd Time Series with Regime Coloring}
    \label{fig:svd-ts-colored}
  \end{subfigure}
  \caption{[[Maybe this should be just a picture of each time series and then have another figure with color after explaining heuristically the reigmes. In (a) the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
    instruction period. Similarly in (b) is a time series of the IPC during the execution of \gcc.}\label{fig:sample-ts}
    \end{figure}

### 2014-03-02 model.tex
adding experimental methods/collection section and updating

\begin{enumerate}
\item \cmark section intro
\item \cmark fix order from simplest to most complex
\item Description of DCE and parameter estimation  
\item Description of auto ARIMA 
\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
\item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
\item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.

@@

While the simplicity of these methods is advantageous when a signal is mostly complexity, i.e., that is the time series exists on the high side of the complexity spectrum, the methods can be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series. Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths. Fortunately, they somewhat parallel each other, i.e., the weakness of Random Walk and \naive are covered by the other primitive method. The situation in which Random Walk fails is when a time series changes drastically on a regular basis. The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step. At every time step Random Walk would preciesly predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$. In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue and error of 0.5 at every step. So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.

@@

For this \cite{davislinearts}

@@

 \begin{figure}
   \centering
     \includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{[[{\color{red}Joshua:Perhaps add a similar graphic of \gcc to show contrast}{\color{green}agreed I will work on this}]]A 3D projection of a delay-coordinate embedding of the trace
 from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
 }
 \label{fig:embedding}
 \end{figure}

>>>>

\begin{enumerate}
\item \cmark section intro
\item \cmark fix order from simplest to most complex
\item Description of DCE and parameter estimation  
\item\cmark Description of auto ARIMA 
\item \cmark Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
\item\cmark Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.

@@

While the simplicity of these methods is advantageous on average when a time series exists on the high side of the complexity spectrum, their simplicity can also be a disadvantage and cause the methods to be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series. Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths. Fortunately, they somewhat parallel each other, i.e., the weakness of Random Walk and \naive are covered by the other primitive method. The situation in which Random Walk fails is when a time series changes drastically on a regular basis. The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step. At every time step Random Walk would precisely predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$. In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue and error of 0.5 at every step. So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.

@@

%For this \cite{davislinearts}

@@

 \begin{figure}
   \centering
\begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%   
  \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colipc3d.png}
    \caption{\gcc}
    \label{fig:gccEmbedding}
  \end{subfigure}%

     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{[[{\color{red}Joshua:Perhaps add a similar graphic of \gcc to show contrast}{\color{green}agreed I will work on this}]]A 3D projection of a delay-coordinate embedding of the trace
 from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
 }
 \label{fig:embedding}
 \end{figure}
